INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-10-13 chapters/background.tex
related projects

\paragraph{BaseLanguage.}
Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}~\cite{ref:BaseLanguage}.
It is a clone of Java implemented using the MPS constructs.
BaseLanguage was developed in the early days of MPS in order to implement MPS itself and also to support the basic set of language-definition DSLs,
Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all the languages in MPS.
The language-definition DSLs, used to define custom languages by their authors, are generated into the BaseLanguage.
Similarly, all the custom DSLs that are meant to be generated into Java choose BaseLanguage as their generation target.
The conversion to textual Java is handled by BaseLanguage without any further manual effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.

@@

The BaseLanguage~\cite{ref:BaseLanguage} is an almost full port of the Java language, extended with MPS-specific features.
It was created manually by JetBrains, and it is still undergoing changes as Java itself is evolving.

@@

% TODO continue from here

\paragraph{ANTLR{\_}MPS.}
\PV{Tenhle projekt byl z tech tri nejmene uzitecny a podle me i nejmene funkcni, tak mozna by se dal v ramci zkracovani vypustit. V dalsim projektu je na nej reference, tak treba ji vyhodit.}
ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
The author of this project is Fabien Campagne.
The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.

The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
It works as follows. We give only a brief high-level overview and omit low-level technical details.
The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
To import a language, the user utilizes this MPS ANTLR language.
The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
There are no editor nor TextGen aspects created neither.

\paragraph{mps-metabnf.}
\PV{Pridal jsem popis projektu}
% TODO - citace na https://github.com/DSLFoundry/mps-metabnf
The mps-metabnf project comes from the DSLFoundry group and also takes on grammar importing.
Even though it is in a very early stage of development, it holds some interesting ideas.
Similarly to ANTLR{\_}MPS, authors of mps-metabnf decided to create an MPS language describing the grammar.
When grammar is being imported into MPS, it is just transformed into the terms of this MPS language.
User is then able to adjust this grammar.
After this, a second step of the import process is needed, that generates the final MPS language out of this adjusted grammar definition.

The key takeaways from the mps-metabnf project are that the intermediate grammar MPS representation, when finished, will offer some powerful means of adjustment.
Using the projectional editor, the user can easily specify information about the code layout, export the target MPS language and, in case further adjustments are needed, come back to the grammar definition, fix any problems and regenerate the language.
The MPS language of the grammar definition can be easily extended with needed features (indenting, line breaks...), which is superior to our approach, because they can be represented using many visual means.

This approach also has some downsides.
Because MPS doesn't offer any tools that would allow easy transformation from the MPS grammar \PP{To je konkretni gramatika zadefinovana pomoci MPS jazyka pro ANTLR} to the MPS language, the second step of the import process hasn't been implemented yet and will be probably very problematic.
Another downside might be the need for a double transformation implementation -- first one being the import of the text grammar to the MPS grammar definition.

From another point of view -- for each additional feature, that we decide to add to the grammar notation, we need to implement a corresponding transformation.
This subsequently leads to duplicating the projectional editor that is already present in MPS.
This means that there needs to be a line drawn between which features are added and which will be left for the user to add to the target language using MPS.

>>>>

\paragraph{BaseLanguage.}
Another important feature of MPS that we need to describe in more detail is \emph{BaseLanguage}~\cite{ref:BASELANG}.
It is a clone of Java implemented using the MPS constructs.
BaseLanguage was developed in the early days of MPS in order to implement MPS itself and also to support the basic set of language-definition DSLs,
Although BaseLanguage is syntactically almost identical to Java, it is edited in a projectional editor, just like all the languages in MPS.
The language-definition DSLs, used to define custom languages by their authors, are generated into the BaseLanguage.
Similarly, all the custom DSLs that are meant to be generated into Java choose BaseLanguage as their generation target.
The conversion to textual Java is handled by BaseLanguage without any further manual effort, since BaseLanguage has a TextGen aspect defined, which translates BaseLanguage code into textual Java sources.

@@

The BaseLanguage~\cite{ref:BASELANG} is an almost full port of the Java language, extended with MPS-specific features.
It was created manually by JetBrains, and it is still undergoing changes as Java itself is evolving.

@@

\paragraph{ANTLR{\_}MPS.}
The ANTLR{\_}MPS project~\cite{ref:ANTLR2MPS} also uses the ANTLR v4 grammar notation, but otherwise works quite differently from PE4MPS.
We provide only a brief mention here because this project is in an early stage of development.
Its author created an ANTLRv4 MPS language, which captures the syntax of ANTLRv4 notation using MPS concepts.
Given some language as input, the textual grammar of the language is imported automatically into MPS, taking the form of the ANTLRv4 MPS language.
The next step would be to create a new MPS language from the grammar definition in MPS, but it is not implemented yet.

However, the project has also other important limitations.
The tool does not generate any parent-child relationships in the Structure aspect, and it completely neglects the Editor and TextGen aspects.
In particular, it does not support creating of the projectional editor at all.

\paragraph{mps-metabnf.}
The mps-metabnf tool~\cite{ref:MPSMETABNF}, implemented by members of the DSLFoundry group, builds upon similar ideas as the ANTLR{\_}MPS project described above.
Authors of this project, too, created an MPS language that can be used to capture ANTLR grammars.
Every imported grammar is represented using this ANTLRv4 MPS language.
A user is then able to adjust the grammar as he wishes, specifying the code layout in the projectional editor.
In the last step, the target MPS language is derived from the adjusted grammar definition.
Many iterations of this process, involving adjustments of the grammar followed by recreation of the target language, can be performed.

The main limitation of this approach is the need to perform the last step manually, because MPS currently does not provide any support for automated transformation of a grammar captured by the ANTLRv4 MPS language into the actual target language.

### 2016-10-13 paper.tex
text about pe4mps
related projects

@@

@@

\bibitem{ref:BaseLanguage} MPS BaseLanguage, \url{https://confluence.jetbrains.com/display/MPSD34/Base+Language}

@@

\bibitem{ref:PE4MPS} PE4MPS project, \url{https://github.com/mar9000/pe4mps}

\bibitem{ref:PE} PE project, \url{https://github.com/mar9000/pe}

\bibitem{ref:ANTLR2MPS} ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR{\_}MPS}

\bibitem{ref:mpsmetabnf} mps-metabnf \url{https://github.com/DSLFoundry/mps-metabnf}

>>>>

\bibitem{ref:ANTLRBOOK} T. Parr. The Definitive ANTLR 4 Reference. Pragmatic Bookshelf, 2013.

@@

\bibitem{ref:ANTLR2MPS} F. Campagne. The ANTRL{\_}MPS project, \url{https://github.com/CampagneLaboratory/ANTLR_MPS}

\bibitem{ref:MPSMETABNF} DSLFoundry. The mps-metabnf project, \url{https://github.com/DSLFoundry/mps-metabnf}

@@

\bibitem{ref:PE4MPS} M. Lombardo. The PE4MPS project, \url{https://github.com/mar9000/pe4mps}

\bibitem{ref:PE} M. Lombardo. PE grammars, \url{https://github.com/mar9000/pe}

\bibitem{ref:BASELANG} MPS BaseLanguage, \url{https://confluence.jetbrains.com/display/MPSD34/Base+Language}

@@
